HyperLogLog: Exponentially Bad in Adversarial Settings

### 2021-05-05 docs/HLL_exp_bad/main.tex
updates after feedback from Mia and Anu

\begin{abstract}
Computing the count of distinct elements in large data sets is a common task but naive approaches are memory-expensive. The HyperLogLog (HLL) algorithm (Flajolet \emph{et al.}, 2007) estimates a data set's cardinality while using significantly less memory than a naive approach, at the cost of some accuracy. This trade-off makes the HLL algorithm very attractive for a wide range of applications such as database management and network monitoring, where an exact count may not be needed. The HLL algorithm and variants of it are implemented in systems such as Redis and Google Big Query. Recently, the HLL algorithm has started to be proposed for use in scenarios where the inputs may be adversarially generated, for example counting social network users or detection of network scanning attacks. This prompts an examination of the performance of the HLL algorithm in the face of adversarial inputs. We show that in such a setting, the HLL algorithm's estimate of cardinality can be exponentially bad: when an adversary has access to the internals of the HLL algorithm and has some flexibility in choosing what inputs will be recorded, it can manipulate the cardinality estimate to be exponentially smaller than the true cardinality. We study both the original HLL algorithm and a more modern version of it (Ertl, 2017) that is used in Redis. We present experimental results confirming our theoretical analysis. Finally, we consider attack prevention: we show to modify HLL in a simple way that provably prevents cardinality estimate manipulation attacks.
\end{abstract}

@@

In this section we present a formal security analysis of HLL, focussing on how to augment the original HLL design with cryptographic mechanisms and thereby provably prevent attacks.

@@

We first begin by explaining the high-level idea of our approach. We allow an adversary ${\mathcal{A}}$ to interact with an HLL via various oracles that capture its capabilities in each of the three scenarios S2, S3 and S4: it may have an \texttt{insert($\cdot$)} oracle allowing it to add items to the HLL sketch, a \texttt{read} oracle that returns the HLL cardinality estimate, and a \texttt{corrupt} that returns all the HLL bucket contents (but not any secret key $K$); in S4 it has all three oracles. In addition, in all 3 scenarios, ${\mathcal{A}}$ has access to an \texttt{init($\cdot$)} oracle by which it sets the HLL parameters $n$, $\ell$, $\alpha_n$ used to initialise the HLL sketch. We assume ${\mathcal{A}}$ always makes exactly one \texttt{init($\cdot$)} query and that it is its first oracle call.

@@

In the ideal world ($b=1$), ${\mathcal{A}}$ is replaced by a simulator ${\mathcal{S}}$ with blackbox oracle access to ${\mathcal{A}}$ (that is, ${\mathcal{S}}$ can run ${\mathcal{A}}$ in a blackbox manner, receiving its oracle queries and providing responses to ${\mathcal{A}}$ as it sees fit). In particular, ${\mathcal{S}}$ receives parameters $n$, $\ell$, $\alpha_n$ from ${\mathcal{A}}$'s \texttt{init($\cdot$)} call. Finally, ${\mathcal{S}}$ receives $\textsf{out}_{\mathcal{A}}$ and completes its execution by producing its own output $\textsf{out}_{\mathcal{S}}$.

@@

The four oracles that  ${\mathcal{A}}$ can interact with behave as follows in the real world:
\begin{itemize}
\item \texttt{init($\cdot$)}: this oracle receives as input the HLL-$F$ parameters ($n$, $\ell$, $\alpha_n$) and performs initialisation of the HLL-$F$ sketch, including generating the array of buckets $M[j]$ and initialising all the entries to $0$, and selecting a key $K \leftarrow_{\$} \{0,1\}^k$ for keyed function $F: \{0,1\}^k \times {{\mathcal{D}}} \rightarrow  \{0,1\}^{n+\ell}$. This oracle has no output. It is available in all of S2, S3, S4. We assume it is only called once as ${\mathcal{A}}$'s first oracle query.
\item \texttt{insert($\cdot$)}: on input $x \in {{\mathcal{D}}}$, the oracle implements HLL-$F$ by setting $y = F_K(x)$ and following the procedure in Figure~\ref{fig:hll} to insert $x$ into the HLL sketch (that is, it uses the first $m$ bits of $y$ to determine a bucket index $j$ and the next $\ell$ bits of $y$ to determine a bit-string $w$; then $\rho(w)$ is used to update the bucket at index $j$ via $M[j]:= \max(M[j],\rho(w))$). This oracle has no output. It is available in all of S2, S3, S4. Without loss of generality (since HLL and variants of it we consider here have a deterministic insertion procedure), we assume all queries $x$ made by ${\mathcal{A}}$ to \texttt{insert($\cdot$)} are distinct.
\item \texttt{read}: this oracle has no input and returns the HLL-$F$ cardinality estimate as per Figure~\ref{fig:hll}, i.e.\ it computes a raw cardinality estimate via $E := \alpha_mm^2\cdot(\sum_{j=1}^m2^{-M[j]})^{-1}$ and then performs corrections for small and large values. We allow multiple queries to this oracle and make it available to ${\mathcal{A}}$ in S3 and S4, but not S2.
\item \texttt{corrupt}: this oracle has no input and returns the vector $(M[j])_{j=1}^{m}$ of HLL bucket contents (but not any secret key $K$). This oracle is available only in S4. We allow the adversary to query \texttt{corrupt} as often as it wishes during its attack. Our description of S4 in Section~\ref{sec:attacks} limited the adversary to at most one such query; here we allow multiple queries because doing so does not increase the difficulty of achieving security.  
\end{itemize}

@@

We leave the exploration of the adaptation of our simulation-based security definitions to capture confidentiality notions for more restrictred APIs to future work.

@@

It is natural to wonder whether a PRF is really required in place of hash function $h$ in order to make HLL secure, or whether a weaker cryptographic primitive would suffice. For example, since it has (roughly) uniform outputs, perhaps using a universal hash function (UHF) $h_K(\cdot)$ would be sufficient. In Appendix~\ref{app:UHF} we show that, in general, UHFs do not suffice by providing a key recovery attack against HLL when $h$ is replaced by a specific UHF based on polynomial evaluation over a finite field.

@@

PRF adversary ${{\mathcal{B}}}$ against $F$ is built as follows. Let $d$ denote the hidden bit determining whether ${{\mathcal{B}}}$ receives real responses of the form $F_K(x)$ where $K$ is a uniformly random key ($d=0$) or responses from a truly random function ($d=1$); ${{\mathcal{B}}}$ provides a simulation of either the real or random world for distinguisher ${\mathcal{E}}$ as follows. ${{\mathcal{B}}}$ runs ${{\mathcal{A}}}$ and answers each of ${{\mathcal{A}}}$'s queries as follows:
\begin{itemize}
\item \texttt{init($\cdot$)}: ${{\mathcal{B}}}$ receives parameters $n$, $\ell$, $\alpha_m$ and initialises an array $M[j]$ in which each of the $2^n$ entries is set to 0. 
\item \texttt{insert($\cdot$)}: on input $x \in {{\mathcal{D}}}$,  ${{\mathcal{B}}}$ makes a query on $x$ to its real-or-random oracle and receives value $y$; ${{\mathcal{B}}}$ uses $y$ to update the HLL sketch as in the regular HLL algorithm (shown in Figure~\ref{fig:hll}). That is, it uses the first $m$ bits of $y$ to determine a bucket index $j$ and the next $\ell$ bits of $y$ to determine a bit-string $w$; then $\rho(w)$ is used to update the bucket at index $j$ via $M[j]:= \max(M[j],\rho(w))$.
\item \texttt{read}: ${{\mathcal{B}}}$ uses its knowledge of the buckets' contents to compute and return the HLL cardinality estimate as per Figure~\ref{fig:hll}. 
\item \texttt{corrupt}: ${{\mathcal{B}}}$ returns the vector $(M[j])_{j=1}^{m}$. 
\end{itemize}
When ${\mathcal{A}}$ returns its final output $\textsf{out}_{\mathcal{S}}$, ${{\mathcal{B}}}$ forwards this output to ${{\mathcal{E}}}$. Finally, ${{\mathcal{E}}}$ outputs a bit $b'$. Then ${{\mathcal{B}}}$ outputs its own bit $d'=b'$.

>>>>

\begin{abstract}
Computing the count of distinct elements in large data sets is a common task but naive approaches are memory-expensive. The HyperLogLog (HLL) algorithm (Flajolet \emph{et al.}, 2007) estimates a data set's cardinality while using significantly less memory than a naive approach, at the cost of some accuracy. This trade-off makes the HLL algorithm very attractive for a wide range of applications such as database management and network monitoring, where an exact count may not be needed. The HLL algorithm and variants of it are implemented in systems such as Redis and Google Big Query. Recently, the HLL algorithm has started to be proposed for use in scenarios where the inputs may be adversarially generated, for example counting social network users or detection of network scanning attacks. This prompts an examination of the performance of the HLL algorithm in the face of adversarial inputs. We show that in such a setting, the HLL algorithm's estimate of cardinality can be exponentially bad: when an adversary has access to the internals of the HLL algorithm and has some flexibility in choosing what inputs will be recorded, it can manipulate the cardinality estimate to be exponentially smaller than the true cardinality. We study both the original HLL algorithm and a more modern version of it (Ertl, 2017) that is used in Redis. We present experimental results confirming our theoretical analysis. Finally, we consider attack prevention: we show how to modify HLL in a simple way that provably prevents cardinality estimate manipulation attacks.
\end{abstract}

@@

In this section we present a formal security analysis of HLL, focussing on how to augment the original HLL design with cryptographic mechanisms and thereby provably prevent attacks. We focus on the case of an initially empty HLL sketch, giving the adversary full control over which items are inserted into it.

@@

We first begin by explaining the high-level idea of our approach. We allow an adversary ${\mathcal{A}}$ to interact with an HLL via various oracles that capture its capabilities in each of the three scenarios S2, S3 and S4: it may have an \texttt{insert($\cdot$)} oracle allowing it to add items to the HLL sketch, a \texttt{read} oracle that returns the HLL cardinality estimate, and a \texttt{corrupt} that returns all the HLL bucket contents (but not any secret key $K$); in S4 it has all three oracles. In addition, in all 3 scenarios, ${\mathcal{A}}$ has access to an \texttt{init($\cdot$)} oracle by which it sets the HLL parameters $n$, $m=2^n$, $\ell$, $\alpha_m$ used to initialise the HLL sketch. We assume ${\mathcal{A}}$ always makes exactly one \texttt{init($\cdot$)} query and that it is its first oracle call.

@@

In the ideal world ($b=1$), ${\mathcal{A}}$ is replaced by a simulator ${\mathcal{S}}$ with blackbox oracle access to ${\mathcal{A}}$ (that is, ${\mathcal{S}}$ can run ${\mathcal{A}}$ in a blackbox manner, receiving its oracle queries and providing responses to ${\mathcal{A}}$ as it sees fit). In particular, ${\mathcal{S}}$ receives parameters $n$, $\ell$, $\alpha_m$ from ${\mathcal{A}}$'s \texttt{init($\cdot$)} call. Finally, ${\mathcal{S}}$ receives $\textsf{out}_{\mathcal{A}}$ and completes its execution by producing its own output $\textsf{out}_{\mathcal{S}}$.

@@

The four oracles that  ${\mathcal{A}}$ can interact with behave as follows in the real world:
\begin{itemize}
\item \texttt{init($\cdot$)}: this oracle receives as input the HLL-$F$ parameters ($n$, $\ell$, $\alpha_m$) and performs initialisation of the HLL-$F$ sketch, including generating the array of $m=2^n$ buckets $M[j]$ and initialising all the entries to $0$, and selecting a key $K \leftarrow_{\$} \{0,1\}^k$ for keyed function $F: \{0,1\}^k \times {{\mathcal{D}}} \rightarrow  \{0,1\}^{n+\ell}$. This oracle has no output. It is available in all of S2, S3, S4. We assume it is only called once as ${\mathcal{A}}$'s first oracle query.
\item \texttt{insert($\cdot$)}: on input $x \in {{\mathcal{D}}}$, the oracle implements HLL-$F$ by setting $y = F_K(x)$ and following the procedure in Figure~\ref{fig:hll} to insert $x$ into the HLL sketch (that is, it uses the first $n$ bits of $y$ to determine a bucket index $j$ and the next $\ell$ bits of $y$ to determine a bit-string $w$; then $\rho(w)$ is used to update the bucket at index $j$ via $M[j]:= \max(M[j],\rho(w))$). This oracle has no output. It is available in all of S2, S3, S4. Since HLL and variants of it we consider here have the property that reinserting an item $x$ does not change the internal state of the HLL sketch (nor the cardinality estimate computed from it) we assume without loss of generality that all queries $x$ made by ${\mathcal{A}}$ to \texttt{insert($\cdot$)} are distinct.
\item \texttt{read}: this oracle has no input and returns the HLL-$F$ cardinality estimate as per Figure~\ref{fig:hll}, i.e.\ it computes a raw cardinality estimate via $E := \alpha_mm^2\cdot(\sum_{j=1}^m2^{-M[j]})^{-1}$ and then performs corrections for small and large values. We allow multiple queries to this oracle and make it available to ${\mathcal{A}}$ in S3 and S4, but not S2.
\item \texttt{corrupt}: this oracle has no input and returns the vector $(M[j])_{j=1}^{m}$ of HLL bucket contents (but not any secret key $K$). This oracle is available only in S4. We allow the adversary to query \texttt{corrupt} as often as it wishes during its attack. Our description of S4 in Section~\ref{sec:attacks} limited the adversary to at most one such query; here we allow multiple queries because doing so does not increase the difficulty of achieving security.  
\end{itemize}

@@

We leave the exploration of the adaptation of our simulation-based security definitions to capture confidentiality notions for more restricted APIs to future work.

@@

It is natural to wonder whether a PRF is really required in place of hash function $h$ in order to make HLL secure, or whether a weaker cryptographic primitive would suffice. For example, since it has (roughly) uniform outputs, perhaps using a universal hash function (UHF) $h_K(\cdot)$ would be sufficient. In Appendix~\ref{sec:UHF} we show that, in general, UHFs do not suffice by providing a key recovery attack against HLL when $h$ is replaced by a specific UHF based on polynomial evaluation over a finite field.

@@

PRF adversary ${{\mathcal{B}}}$ against $F$ is built as follows. Let $d$ denote the hidden bit determining whether ${{\mathcal{B}}}$ receives real responses of the form $F_K(x)$ where $K$ is a uniformly random key ($d=0$) or responses from a truly random function ($d=1$); ${{\mathcal{B}}}$ provides a simulation of either the real or ideal world for distinguisher ${\mathcal{E}}$ as follows. ${{\mathcal{B}}}$ runs ${{\mathcal{A}}}$ and answers each of ${{\mathcal{A}}}$'s queries as follows:
\begin{itemize}
\item \texttt{init($\cdot$)}: ${{\mathcal{B}}}$ receives parameters $n$, $\ell$, $\alpha_m$, sets $m=2^n$ and initialises an array $M[j]$ of size $m$ in which each of the entries is set to 0. 
\item \texttt{insert($\cdot$)}: on input $x \in {{\mathcal{D}}}$,  ${{\mathcal{B}}}$ makes a query on $x$ to its real-or-random oracle and receives value $y$; ${{\mathcal{B}}}$ uses $y$ to update the HLL sketch as in the regular HLL algorithm (shown in Figure~\ref{fig:hll}). That is, it uses the first $n$ bits of $y$ to determine a bucket index $j$ and the next $\ell$ bits of $y$ to determine a bit-string $w$; then $\rho(w)$ is used to update the bucket at index $j$ via $M[j]:= \max(M[j],\rho(w))$.
\item \texttt{read}: ${{\mathcal{B}}}$ uses its knowledge of the buckets' contents to compute and return the HLL cardinality estimate as per Figure~\ref{fig:hll}. 
\item \texttt{corrupt}: ${{\mathcal{B}}}$ returns the vector $(M[j])_{j=1}^{m}$. 
\end{itemize}
When ${\mathcal{A}}$ returns its final output $\textsf{out}_{\mathcal{A}}$, ${{\mathcal{B}}}$ forwards this output to ${{\mathcal{E}}}$. Finally, ${{\mathcal{E}}}$ outputs a bit $b'$. Then ${{\mathcal{B}}}$ outputs its own bit $d'=b'$.
